---
title: "SBD3_1_Assignment"
format: html
---

# Libraries

```{r}
library(dlookr)
library(caret)
library(rsample)
```

# Setup

```{r}
rm(list=ls())
options(scipen=999)
getwd()
set.seed(123)
```

# Loading Data

```{r}
load("data_wage.RData")
head(data)
```

# Data Checks

```{r}
summary(data)
```

```{r}
diagnose(data)
# No Missing Values in the Dataset
# All the numeric variables have only 2 values (0 and 1) except the target variable wage
```

```{r}
str(data)

# All the non numeric variables are factors with levels.
```

```{r}
colnames(data)
# Some of the Columns do not really add any value when considering the other variables in the dataset
# For example:
#
# "Activities_None.of.these.activities.are.an.important.part.of.my.role.at.work"
# "Notebooks_None" 
# "cloud_I.have.not.used.any.cloud.providers" 
# "ML_framework_None"
# "Visualization_None"
# "explainability.model_None.I.do.not.use.these.model.explanation.techniques"
```

## Numeric

### Target Variable

```{r}
summary(data$wage)
sd(data$wage)
# right skewed distribution with a long tail. The mean is higher than the median, which indicates that there are some high wage values that are pulling the average up. The standard deviation is also quite large, which suggests that there is a lot of variability in wages among the individuals in the dataset.
```

```{r}
boxplot(data$wage, horizontal = TRUE, main = "Boxplot of Wage", xlab = "Wage")
hist(data$wage, breaks = 50, main = "Histogram of Wage", xlab = "Wage")
```

```{r}
# Log Wage
hist(log(data$wage[data$wage > 0]), breaks = 50, main = "Histogram of log(wage)")
boxplot(log(data$wage[data$wage > 0]), horizontal = TRUE)
```

```{r}
# 0 in Target variable "wage"
sum(data$wage == 0)
prop.table(table(data$wage == 0))
data_no_0_wage <- data[data$wage > 0, ]
```

```{r}
boxplot(data_no_0_wage$wage, horizontal = TRUE, main = "Boxplot of Wage", xlab = "Wage")
hist(data_no_0_wage$wage, breaks = 50, main = "Histogram of Wage", xlab = "Wage")
```

### Features

```{r}
#Frequency
print(table(data$Notebooks_None))
  
#Proportions 
print(prop.table(table(data$Notebooks_None))) 
```

## Factors

```{r}
#Levels
print(levels(data$gender))
  
#Frequency
print(table(data$gender))
  
#Proportions
print(round(prop.table(table(data$gender)), 2))

#Ordered
print(is.ordered(data$gender))
# All the factors are unordered

#Variable = ML_atWork
#At first, we thought that the response "I do not know" would not be useful. However, it could be that interns genuinely do not know this. Additionally, there are many 
#people who are not working—what did they select? Possibly this answer as well. There should be an answer option such as "I do not work".


#Variable = Do.you.consider.yourself.to.be.a.data.scientist
#The responses are purely subjective, which is why we will probably remove this variable.
```

# Data Cleaning

## Order Factors

```{r}
# The following (factor) variables were treated as ordered:
# - Age
# - education
# - years_experience
# - How.long.have.you.been.writing.code.to.analyze.data.
# - percent_actively.coding
# - For.how.many.years.have.you.used.machine.learning.meth
# ods..at.work.or.in.school..
# - Do.you.consider.yourself.to.be.a.data.scientist.
# - ML_atWork
```

```{r}
# Gender
levels(data$gender)
data$gender <- factor(data$gender, ordered = FALSE)
```

```{r}
# Age
levels(data$age)
data$age <- factor(
  data$age,
  levels = c(
    "18-21", "22-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-69", "70-79", "80+"
  ),
  ordered = TRUE
)
```

```{r}
# Country
levels(data$country)
data$country <- factor(data$country, ordered = FALSE)
```

```{r}
# Education
levels(data$education)
data$education <- factor(
  data$education,
  levels = c(
    "Some college/university study without earning a bachelor’s degree",
    "Bachelor’s degree",
    "Master’s degree",
    "Professional degree",
    "Doctoral degree"
  ),
  ordered = TRUE
)
# Professional degree is hard to rank here. Maybe we should make this unordered.
```

```{r}
# undergraduate_major
levels(data$undergraduate_major)
data$undergraduate_major <- factor(data$undergraduate_major, ordered = FALSE)
```

```{r}
# job_role 
levels(data$job_role)
data$job_role <- factor(data$job_role, ordered = FALSE)
```

```{r}
# industry
levels(data$industry)
data$industry <- factor(data$industry, ordered = FALSE)
```

```{r}
# years_experience
levels(data$years_experience)
data$years_experience <- factor(
  data$years_experience,
  levels = c(
    "0-1", "1-2", "2-3", "3-4", "4-5", "5-11", "11-15", "15-20", "20-25", "25-30", "30 +"
  ),
  ordered = TRUE
)
```

```{r}
# ML_atwork
levels(data$ML_atwork)
data$ML_atwork <- factor(
  data$ML_atwork,
  levels = c(
    "I do not know",
    "No (we do not use ML methods)",
    "We are exploring ML methods (and may one day put a model into production)",
    "We use ML methods for generating insights (but do not put working models into production)",
    "We recently started using ML methods (i.e., models in production for less than 2 years)",
    "We have well established ML methods (i.e., models in production for more than 2 years)"
    ),
    ordered = FALSE
  )

# We also wanted to order this variable, but we could not agree, as the ordering depends on the perspective taken.
```

```{r}
# Programming_language_used_most_often
levels(data$Programming_language_used_most_often)
data$Programming_language_used_most_often <- factor(data$Programming_language_used_most_often, ordered = FALSE)
```

```{r}
# percent_actively.coding
levels(data$percent_actively.coding)
data$percent_actively.coding <- factor(
  data$percent_actively.coding,
  levels = c(
    "0% of my time", "1% to 25% of my time", "25% to 49% of my time", "50% to 74% of my time", "75% to 99% of my time", "100% of my time"
  ),
  ordered = TRUE
)
```

```{r}
# How.long.have.you.been.writing.code.to.analyze.data.
levels(data$How.long.have.you.been.writing.code.to.analyze.data.)
data$How.long.have.you.been.writing.code.to.analyze.data. <- factor(
  data$How.long.have.you.been.writing.code.to.analyze.data.,
  levels = c(
    "I have never written code and I do not want to learn", 
    "I have never written code but I want to learn", 
    "< 1 year", 
    "1-2 years", 
    "3-5 years", 
    "5-10 years", 
    "10-20 years", 
    "20-30 years", 
    "30-40 years", 
    "40+ years"
  ),
  ordered = TRUE
)
```

```{r}
# For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..
levels(data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..)
data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school.. <- factor(
  data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..,
  levels = c(
    "I have never studied machine learning and I do not plan to", 
    "I have never studied machine learning but plan to learn in the future", 
    "< 1 year", 
    "1-2 years", 
    "2-3 years", 
    "3-4 years", 
    "4-5 years", 
    "5-10 years", 
    "10-15 years", 
    "15-20 years",
    "20+ years"
  ),
  ordered = TRUE
)
```

```{r}
# Do.you.consider.yourself.to.be.a.data.scientist.
levels(data$Do.you.consider.yourself.to.be.a.data.scientist.)
data$Do.you.consider.yourself.to.be.a.data.scientist. <- factor(
  data$Do.you.consider.yourself.to.be.a.data.scientist.,
  levels = c(
    "Definitely not", 
    "Probably not",
    "Maybe",
    "Probably yes",
    "Definitely yes"
  ),
  ordered = TRUE
)
```

## Outliers

```{r}
# Just as a reminder if necessary
```

# Exploratory Data Analysis

```{r}
# Only if needed
```

# Data Preparation

```{r}
data_prepared <- data
# General feature selection:
# We remove "Do you consider yourself to be a data scientist?"
data_prepared$Do.you.consider.yourself.to.be.a.data.scientist. <- NULL
# Remove wage = 0 before creating the log outcome.
data_prepared <- subset(data_prepared, wage > 0)
# Log wage:
# We keep the original wage column for interpretation,but create log_wage as the modelling target.
data_prepared$log_wage <- log(data_prepared$wage)
# Drop unused levels after filtering
data_prepared <- droplevels(data_prepared)
cat("Rows in prepared data:", nrow(data_prepared), "\n")
cat("Columns in prepared data:", ncol(data_prepared), "\n")
summary(data_prepared$log_wage)
```

# General Train Test Split

```{r}
split  <- initial_split(data_prepared, prop = 0.7)
data_train  <- training(split)
data_test   <- testing(split)

hist(data_train$log_wage, main = "Histogram of log_wage in train data", xlab = "log_wage")
hist(data_test$log_wage, main = "Histogram of log_wage in test data", xlab = "log_wage")
```

# Models

## Linear Regression (Benchmark)

```{r}
linear_regression_model <- lm(log_wage ~ . - wage,
                              data = data_train)
```

```{r}
# Print results of the model
summary(linear_regression_model)
```

```{r}
# Make predictions on the test data to evaluate  model's performance
linear_regression_model_predictions <- linear_regression_model |> predict(data_test)
```

```{r}
# Compute the prediction error, RMSE
linear_regression_rmse <- RMSE(linear_regression_model_predictions, data_test$log_wage)
cat("RMSE (Linear Regression, log_wage):", linear_regression_rmse)
```

```{r}
linear_regression_model <- lm(log_wage ~ 
                                age +
                                country +
                                job_role +
                                industry +
                                ML_atwork,
                              data = data_train)
```

```{r}
summary(linear_regression_model)
```

```{r}
# Make predictions on the test data to evaluate model's performance
linear_regression_model_predictions <- linear_regression_model |> predict(data_test)
```

```{r}
# Compute the prediction error, RMSE
linear_regression_rmse <- RMSE(linear_regression_model_predictions, data_test$log_wage)
cat("RMSE (Linear Regression, log_wage):", linear_regression_rmse)
```

## Regression Tree

To keep the comparison fair, we use the same predictors that were already used in the reduced linear regression benchmark above: age + country + job_role + industry + ML_atwork.

Regression-tree decisions:

-   I used `method = "anova"` because the target variable `log_wage` is continuous, so this is a regression tree, not a classification tree.

```{r}
tree_model <- rpart::rpart(
  log_wage ~ age +
    country +
    job_role +
    industry +
    ML_atwork,
  data = data_train,
  method = "anova",
  control = rpart::rpart.control(cp = 0, maxdepth = 3)
)
```

```{r}
# Print the fitted tree object
tree_model
```

```{r, fig.width=18, fig.height=10, out.width="100%"}

if (requireNamespace("rpart.plot", quietly = TRUE)) {
  rpart.plot::rpart.plot(
    tree_model,
    type = 5,
    cex = 0.6,
    tweak = 0.8,
    fallen.leaves = TRUE
  )
} else {
  par(mar = c(2, 2, 2, 2), xpd = NA)
  plot(
    tree_model,
    uniform = TRUE,
    compress = TRUE,
    branch = 0.4,
    margin = 0.05
  )
  text(
    tree_model,
    use.n = TRUE,
    all = TRUE,
    cex = 0.45,
    pretty = 1
  )
}
```

The plot is useful for understanding the general structure of the tree, but it is too crowded to read every category label clearly because variables like country and job_role have many levels.

```{r}

tree_predictions <- predict(tree_model, data_test)
tree_rmse <- RMSE(tree_predictions, data_test$log_wage)

cat("RMSE (Regression Tree, log_wage):", tree_rmse)
```

The tree on log_wage is more appropriate for this dataset because wages are very skewed, and the log transformation makes the target easier to model.an error of about 0.79 in logs means the model can often be off by roughly a factor of about 2, so it is interpretable, but not super precise.

```{r}

head(
  data.frame(
    actual_log_wage = data_test$log_wage,
    predicted_log_wage = tree_predictions,
    actual_wage = data_test$wage,
    predicted_wage = exp(tree_predictions)
  )
)
```

Because this is a regression tree, several observations can receive the same prediction. the tree does reasonably well for some middle-range cases but it struggles more at the extremes, especially very low and very high wages and it also tends to pull predictions toward group averages.

```{r}
# Variable importance from the fitted regression tree.
tree_model$variable.importance
```

The regression tree suggests that wage differences are driven mainly by country, then by job role and age. This makes sense, because salaries differ strongly across countries, and then also by career position and experience-related factors.

### Improved Regression Tree

The first tree is only a basic implementation. So here we try a slightly deeper tree with a small positive complexity parameter to see if it predicts a bit better without becoming too complex.

```{r}
tree_model_improved <- rpart::rpart(
  log_wage ~ age +
    country +
    job_role +
    industry +
    ML_atwork,
  data = data_train,
  method = "anova",
  control = rpart::rpart.control(cp = 0.001, maxdepth = 4)
)
```

```{r}
tree_predictions_improved <- predict(tree_model_improved, data_test)
tree_rmse_improved <- RMSE(tree_predictions_improved, data_test$log_wage)

cat("RMSE (Improved Regression Tree, log_wage):", tree_rmse_improved)
```

The RMSE is slightly lower than for the first tree, so this version seems to perform better on unseen data. This suggests that allowing a bit more structure helped the model capture wage differences better.

```{r}
comparison_log_trees <- data.frame(
  model = c("Baseline regression tree", "Improved regression tree"),
  cp = c(0, 0.001),
  maxdepth = c(3, 4),
  rmse_log_wage = c(tree_rmse, tree_rmse_improved)
)

comparison_log_trees
```

This table makes the comparison easier to see on the same test data. Since the improved tree has the lower RMSE, I would keep this as the better tree version. The difference is not huge but it still performs better.

```{r, fig.width=18, fig.height=10, out.width="100%"}
if (requireNamespace("rpart.plot", quietly = TRUE)) {
  rpart.plot::rpart.plot(
    tree_model_improved,
    type = 5,
    cex = 0.6,
    tweak = 0.8,
    fallen.leaves = TRUE
  )
} else {
  par(mar = c(2, 2, 2, 2), xpd = NA)
  plot(
    tree_model_improved,
    uniform = TRUE,
    compress = TRUE,
    branch = 0.4,
    margin = 0.05
  )
  text(
    tree_model_improved,
    use.n = TRUE,
    all = TRUE,
    cex = 0.45,
    pretty = 1
  )
}
```

The improved tree is still a bit crowded because some predictors have many levels. But it is still useful for seeing the overall split structure and how the model groups respondents.

## Random Forest

```{r}

```

## Gradient Boosting Models

```{r}

```

## XG Boost

```{r}

```

## Neural Network

```{r}

```

# Model Comparison

```{r}

```

# Final Model

```{r}

```

# Our Wage Prediction

```{r}

```
