---
title: "SBD3_1_Assignment"
format: html
---
# Libraries
```{r}
library(dlookr)
library(rsample)
library(caret)
```

# Setup
```{r}
# rm(list=ls())
options(scipen=999)
getwd()
set.seed(123)
```

# Loading Data
```{r}
load("../01_Submit/data_wage.RData")
head(data)
```

# Data Checks
```{r}
summary(data)
levels(data$education)
levels(data$years_experience)
```

```{r}
diagnose(data)
# No Missing Values in the Dataset
# All the numeric variables have only 2 values (0 and 1) except the target variable wage
```

```{r}
str(data)

# All the non numeric variables are factors with levels.
```

```{r}
colnames(data)
# Some of the Columns do not really add any value when considering the other variables in the dataset
# For example:
#
# "Activities_None.of.these.activities.are.an.important.part.of.my.role.at.work"
# "Notebooks_None" 
# "cloud_I.have.not.used.any.cloud.providers" 
# "ML_framework_None"
# "Visualization_None"
# "explainability.model_None.I.do.not.use.these.model.explanation.techniques"
```
## Numeric

### Target Variable
```{r}
summary(data$wage)
sd(data$wage)
# right skewed distribution with a long tail. The mean is higher than the median, which indicates that there are some high wage values that are pulling the average up. The standard deviation is also quite large, which suggests that there is a lot of variability in wages among the individuals in the dataset.
```


```{r}
boxplot(data$wage, horizontal = TRUE, main = "Boxplot of Wage", xlab = "Wage")
hist(data$wage, breaks = 50, main = "Histogram of Wage", xlab = "Wage")
```

```{r}
# Log Wage
hist(log(data$wage[data$wage > 0]), breaks = 50, main = "Histogram of log(wage)")
boxplot(log(data$wage[data$wage > 0]), horizontal = TRUE)
```

```{r}
# 0 in Target variable "wage"
sum(data$wage == 0)
prop.table(table(data$wage == 0))
data_no_0_wage <- data[data$wage > 0, ]
```
```{r}
boxplot(data_no_0_wage$wage, horizontal = TRUE, main = "Boxplot of Wage", xlab = "Wage")
hist(data_no_0_wage$wage, breaks = 50, main = "Histogram of Wage", xlab = "Wage")
```

### Features
```{r}
#Frequency
print(table(data$Notebooks_None))
  
#Proportions 
print(prop.table(table(data$Notebooks_None))) 
```


## Factors

```{r}
#Levels
print(levels(data$gender))
  
#Frequency
print(table(data$gender))
  
#Proportions
print(round(prop.table(table(data$gender)), 2))

#Ordered
print(is.ordered(data$gender))
# All the factors are unordered

#Variable = ML_atWork
#At first, we thought that the response "I do not know" would not be useful. However, it could be that interns genuinely do not know this. Additionally, there are many 
#people who are not working—what did they select? Possibly this answer as well. There should be an answer option such as "I do not work".


#Variable = Do.you.consider.yourself.to.be.a.data.scientist
#The responses are purely subjective, which is why we will probably remove this variable.
```


# Data Cleaning

## Order Factors
```{r}
# The following (factor) variables were treated as ordered:
# - Age
# - education
# - years_experience
# - How.long.have.you.been.writing.code.to.analyze.data.
# - percent_actively.coding
# - For.how.many.years.have.you.used.machine.learning.meth
# ods..at.work.or.in.school..
# - Do.you.consider.yourself.to.be.a.data.scientist.
# - ML_atWork
```

```{r}
# Gender
levels(data$gender)
data$gender <- factor(data$gender, ordered = FALSE)
```
```{r}
# Age
levels(data$age)
data$age <- factor(
  data$age,
  levels = c(
    "18-21", "22-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-69", "70-79", "80+"
  ),
  ordered = TRUE
)
```
```{r}
# Country
levels(data$country)
data$country <- factor(data$country, ordered = FALSE)
table(data$country)
```

```{r}
# Education
levels(data$education)
data$education <- factor(
  data$education,
  levels = c(
    "I prefer not to answer",
    "Some college/university study without earning a bachelor’s degree",
    "Bachelor’s degree",
    "Master’s degree",
    "Professional degree",
    "Doctoral degree"
  ),
  ordered = FALSE
)
# Professional degree is hard to rank here. Maybe we should make this unordered.
```
```{r}
# undergraduate_major
levels(data$undergraduate_major)
data$undergraduate_major <- factor(data$undergraduate_major, ordered = FALSE)
```
```{r}
# job_role 
levels(data$job_role)
data$job_role <- factor(data$job_role, ordered = FALSE)
```

```{r}
# industry
levels(data$industry)
data$industry <- factor(data$industry, ordered = FALSE)
```
```{r}
# years_experience
levels(data$years_experience)
data$years_experience <- factor(
  data$years_experience,
  levels = c(
    "0-1", "1-2", "2-3", "3-4", "4-5", "5-11", "11-15", "15-20", "20-25", "25-30", "30 +"
  ),
  ordered = TRUE
)
```
```{r}
# ML_atwork
levels(data$ML_atwork)
data$ML_atwork <- factor(
  data$ML_atwork,
  levels = c(
    "I do not know",
    "No (we do not use ML methods)",
    "We are exploring ML methods (and may one day put a model into production)",
    "We use ML methods for generating insights (but do not put working models into production)",
    "We recently started using ML methods (i.e., models in production for less than 2 years)",
    "We have well established ML methods (i.e., models in production for more than 2 years)"
    ),
    ordered = FALSE
  )

# We also wanted to order this variable, but we could not agree, as the ordering depends on the perspective taken.
```
```{r}
# Programming_language_used_most_often
levels(data$Programming_language_used_most_often)
data$Programming_language_used_most_often <- factor(data$Programming_language_used_most_often, ordered = FALSE)
```
```{r}
# percent_actively.coding
levels(data$percent_actively.coding)
data$percent_actively.coding <- factor(
  data$percent_actively.coding,
  levels = c(
    "0% of my time", "1% to 25% of my time", "25% to 49% of my time", "50% to 74% of my time", "75% to 99% of my time", "100% of my time"
  ),
  ordered = TRUE
)
```
```{r}
# How.long.have.you.been.writing.code.to.analyze.data.
levels(data$How.long.have.you.been.writing.code.to.analyze.data.)
data$How.long.have.you.been.writing.code.to.analyze.data. <- factor(
  data$How.long.have.you.been.writing.code.to.analyze.data.,
  levels = c(
    "I have never written code and I do not want to learn", 
    "I have never written code but I want to learn", 
    "< 1 year", 
    "1-2 years", 
    "3-5 years", 
    "5-10 years", 
    "10-20 years", 
    "20-30 years", 
    "30-40 years", 
    "40+ years"
  ),
  ordered = TRUE
)
```
```{r}
# For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..
levels(data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..)
data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school.. <- factor(
  data$For.how.many.years.have.you.used.machine.learning.methods..at.work.or.in.school..,
  levels = c(
    "I have never studied machine learning and I do not plan to", 
    "I have never studied machine learning but plan to learn in the future", 
    "< 1 year", 
    "1-2 years", 
    "2-3 years", 
    "3-4 years", 
    "4-5 years", 
    "5-10 years", 
    "10-15 years", 
    "15-20 years",
    "20+ years"
  ),
  ordered = TRUE
)
```
```{r}
# Do.you.consider.yourself.to.be.a.data.scientist.
levels(data$Do.you.consider.yourself.to.be.a.data.scientist.)
data$Do.you.consider.yourself.to.be.a.data.scientist. <- factor(
  data$Do.you.consider.yourself.to.be.a.data.scientist.,
  levels = c(
    "Definitely not", 
    "Probably not",
    "Maybe",
    "Probably yes",
    "Definitely yes"
  ),
  ordered = TRUE
)
```
## Outliers
```{r}
# Just as a reminder if necessary
```

# Exploratory Data Analysis
## Correlation
```{r}
# Correlation Matrix for numeric variables
cor_matrix <- cor(data[, sapply(data, is.numeric)])  

# Correlation Matrix Plot
corrplot(cor_matrix, method = "color", tl.pos = 'n')
```
```{r}
# Higher correlations only (Threshold: > 0.5 or < -0.5)
high_corr <- which(
  abs(cor_matrix) > 0.5 &
  upper.tri(cor_matrix),
  arr.ind = TRUE
)

high_corr <- data.frame(
  Var1 = rownames(cor_matrix)[high_corr[,1]],
  Var2 = colnames(cor_matrix)[high_corr[,2]],
  Correlation = cor_matrix[high_corr]
)

high_corr <- high_corr[
  order(-abs(high_corr$Correlation)),
]

print(high_corr)
```

# Data Preparation
```{r}
data_clean <- data

# Reducing the amount of levels in the country variable by lumping together the least frequent ones into "Other".
# This is for Tree Models that can not handle more than 53 levels in factors.
# We keep the top 52 countries(52 + Other = 53)
data_clean$country <- fct_lump(data_clean$country, n = 52)

# Removing subjective variable
data_clean$Do.you.consider.yourself.to.be.a.data.scientist. <- NULL
```

# General Train Test Split
```{r}
#Split the Data
split <- initial_split(data_clean, prop = 0.7)

data_train <- training(split)
data_test  <- testing(split)

# Different Data sets to use and compare
# Version A: keep wage = 0
train_all <- data_train
test_all  <- data_test

# Version B: remove wage = 0
train_pos <- train_full[train_full$wage > 0, ]
test_pos  <- test_full[test_full$wage > 0, ]

# Version C: log wage, based on positive wages only
train_log <- train_pos
test_log  <- test_pos
train_log$log_wage <- log(train_log$wage)
test_log$log_wage  <- log(test_log$wage)

# Check the distribution in the Target Variable for all 3 Datasets
# Version A:
hist(train_all$wage) 
hist(test_all$wage) 

# Version B:
hist(train_pos$wage)
hist(test_pos$wage)

# Version C:
hist(train_log$wage)
hist(test_log$wage)
```


# Models

## Linear Regression (Benchmark)
```{r}
# Model 1: wage (all)
set.seed(123)
lm_all <- lm(wage ~ ., data = train_all)

# Model 2: wage (no zeros)
set.seed(123)
lm_pos <- lm(wage ~ ., data = train_pos)

# Model 3: log(wage)
set.seed(123)
lm_log <- lm(log_wage ~ . - wage, data = train_log)
```
```{r}
# Predictions
pred_all <- predict(lm_all, newdata = test_all)

pred_pos <- predict(lm_pos, newdata = test_pos)

pred_log <- predict(lm_log, newdata = test_log)
pred_log_wage <- exp(pred_log)  # back-transform
```

```{r}
results_lm <- data.frame(
  Model = c("LM - wage (all)", "LM - wage (no zeros)", "LM - log(wage)"),
  RMSE = c(
    RMSE(pred_all, test_all$wage),
    RMSE(pred_pos, test_pos$wage),
    RMSE(pred_log_wage, test_pos$wage)
  ),
  MAE = c(
    MAE(pred_all, test_all$wage),
    MAE(pred_pos, test_pos$wage),
    MAE(pred_log_wage, test_pos$wage)
  ),
  R2 = c(
    R2(pred_all, test_all$wage),
    R2(pred_pos, test_pos$wage),
    R2(pred_log_wage, test_pos$wage)
  )
)

results_lm
```
## Decision Tree
```{r}
library(rpart)
library(rpart.plot)
```
```{r}
# Model 1: wage (all)
set.seed(123)
tree_all <- rpart(
  wage ~ .,
  data = train_all,
  method = "anova"
)

# Model 2: wage (no zeros)
set.seed(123)
tree_pos <- rpart(
  wage ~ .,
  data = train_pos,
  method = "anova"
)

# Model 3: log(wage)
set.seed(123)
tree_log <- rpart(
  log_wage ~ . - wage,
  data = train_log,
  method = "anova"
)
```
```{r}
pred_tree_all <- predict(tree_all, newdata = test_all)

pred_tree_pos <- predict(tree_pos, newdata = test_pos)

pred_tree_log <- predict(tree_log, newdata = test_log)
pred_tree_log_wage <- exp(pred_tree_log)
```
```{r}
results_tree <- data.frame(
  Model = c("Tree - wage (all)", "Tree - wage (no zeros)", "Tree - log(wage)"),
  RMSE = c(
    RMSE(pred_tree_all, test_all$wage),
    RMSE(pred_tree_pos, test_pos$wage),
    RMSE(pred_tree_log_wage, test_pos$wage)
  ),
  MAE = c(
    MAE(pred_tree_all, test_all$wage),
    MAE(pred_tree_pos, test_pos$wage),
    MAE(pred_tree_log_wage, test_pos$wage)
  ),
  R2 = c(
    R2(pred_tree_all, test_all$wage),
    R2(pred_tree_pos, test_pos$wage),
    R2(pred_tree_log_wage, test_pos$wage)
  )
)

results_tree
```

## Random Forest
```{r}
library(randomForest)
```
```{r}
# Model 1: wage (all)
set.seed(123)
rf_all <- randomForest(
  wage ~ .,
  data = train_all,
  ntree = 300,
  importance = TRUE
)

# Model 2: wage (no zeros)
set.seed(123)
rf_pos <- randomForest(
  wage ~ .,
  data = train_pos,
  ntree = 300,
  importance = TRUE
)

# Model 3: log(wage)
set.seed(123)
rf_log <- randomForest(
  log_wage ~ . - wage,
  data = train_log,
  ntree = 300,
  importance = TRUE
)
```
```{r}
pred_rf_all <- predict(rf_all, newdata = test_all)

pred_rf_pos <- predict(rf_pos, newdata = test_pos)

pred_rf_log <- predict(rf_log, newdata = test_log)
pred_rf_log_wage <- exp(pred_rf_log)
```
```{r}
results_rf <- data.frame(
  Model = c("RF - wage (all)", "RF - wage (no zeros)", "RF - log(wage)"),
  RMSE = c(
    RMSE(pred_rf_all, test_all$wage),
    RMSE(pred_rf_pos, test_pos$wage),
    RMSE(pred_rf_log_wage, test_pos$wage)
  ),
  MAE = c(
    MAE(pred_rf_all, test_all$wage),
    MAE(pred_rf_pos, test_pos$wage),
    MAE(pred_rf_log_wage, test_pos$wage)
  ),
  R2 = c(
    R2(pred_rf_all, test_all$wage),
    R2(pred_rf_pos, test_pos$wage),
    R2(pred_rf_log_wage, test_pos$wage)
  )
)

results_rf
```

```{r}
importance(rf_all)
varImpPlot(
  rf_all,
  n.var = 15,
  cex = 0.7,
  main = "Top 15 Important Variables"
)

importance_df <- data.frame(
  Variable = rownames(importance(rf_all)),
  IncMSE = importance(rf_all)[, "%IncMSE"],
  NodePurity = importance(rf_all)[, "IncNodePurity"]
)

rownames(importance_df) <- NULL

top_mse <- importance_df[
  order(-importance_df$IncMSE),
]

top_mse <- head(top_mse, 15)

top_purity <- importance_df[
  order(-importance_df$NodePurity),
]

top_purity <- head(top_purity, 15)

par(mar = c(5, 25, 4, 2))

barplot(
  rev(top_mse$IncMSE),
  names.arg = rev(top_mse$Variable),
  horiz = TRUE,
  las = 1,
  cex.names = 0.7,
  main = "Top 15 Variables by % Increase in MSE",
  xlab = "% Increase in MSE"
)

par(mar = c(5, 25, 4, 2))

barplot(
  rev(top_purity$NodePurity),
  names.arg = rev(top_purity$Variable),
  horiz = TRUE,
  las = 1,
  cex.names = 0.7,
  main = "Top 15 Variables by Node Purity",
  xlab = "Increase in Node Purity"
)
```

## Gradient Boosting Models
```{r}
library(gbm)
```
```{r}
# Model 1: wage (all)
set.seed(123)
gbm_all <- gbm(
  formula = wage ~ .,
  data = train_all,
  distribution = "gaussian",
  n.trees = 300,
  interaction.depth = 4,
  shrinkage = 0.03,
  verbose = FALSE
)

# Model 2: wage (no zeros)
set.seed(123)
gbm_pos <- gbm(
  wage ~ .,
  data = train_pos,
  distribution = "gaussian",
  n.trees = 300,
  interaction.depth = 4,
  shrinkage = 0.03,
  verbose = FALSE
)

# Model 3: log(wage)
set.seed(123)
gbm_log <- gbm(
  log_wage ~ . - wage,
  data = train_log,
  distribution = "gaussian",
  n.trees = 300,
  interaction.depth = 4,
  shrinkage = 0.03,
  verbose = FALSE
)
```


```{r}
pred_gbm_all <- predict(gbm_all, test_all, n.trees = 300)

pred_gbm_pos <- predict(gbm_pos, test_pos, n.trees = 300)

pred_gbm_log <- predict(gbm_log, test_log, n.trees = 300)
pred_gbm_log_wage <- exp(pred_gbm_log)
```


```{r}
results_gbm <- data.frame(
  Model = c("GBM - wage (all)", "GBM - wage (no zeros)", "GBM - log(wage)"),
  RMSE = c(
    RMSE(pred_gbm_all, test_all$wage),
    RMSE(pred_gbm_pos, test_pos$wage),
    RMSE(pred_gbm_log_wage, test_pos$wage)
  ),
  MAE = c(
    MAE(pred_gbm_all, test_all$wage),
    MAE(pred_gbm_pos, test_pos$wage),
    MAE(pred_gbm_log_wage, test_pos$wage)
  ),
  R2 = c(
    R2(pred_gbm_all, test_all$wage),
    R2(pred_gbm_pos, test_pos$wage),
    R2(pred_gbm_log_wage, test_pos$wage)
  )
)

results_gbm
```

## XG Boost
```{r}
library(xgboost)
```
```{r}
train_xgb_model <- function(train_data, test_data, target, log_model = FALSE) {
  
  # Create model matrices
  x_train <- model.matrix(
    as.formula(paste(target, "~ .")),
    data = train_data
  )[ , -1]
  
  x_test <- model.matrix(
    as.formula(paste(target, "~ .")),
    data = test_data
  )[ , -1]
  
  y_train <- train_data[[target]]
  
  # Train XGBoost
  model <- xgboost(
    data = x_train,
    label = y_train,
    objective = "reg:squarederror",
    nrounds = 300,
    max_depth = 4,
    eta = 0.05,
    subsample = 0.8,
    colsample_bytree = 0.8,
    verbose = 0
  )
  
  # Predict
  pred <- predict(model, x_test)
  
  if (log_model) {
    pred <- exp(pred)
  }
  
  return(list(model = model, pred = pred))
}
```
```{r}
# XGBoost 1: wage all
set.seed(123)
xgb_all <- train_xgb_model(
  train_all,
  test_all,
  target = "wage"
)

# XGBoost 2: wage no zeros
set.seed(123)
xgb_pos <- train_xgb_model(
  train_pos,
  test_pos,
  target = "wage"
)

# XGBoost 3: log wage
set.seed(123)
xgb_log <- train_xgb_model(
  train_log[, !(names(train_log) %in% "wage")],
  test_log[, !(names(test_log) %in% "wage")],
  target = "log_wage",
  log_model = TRUE
)
```
```{r}
results_xgb <- data.frame(
  Model = c("XGB - wage (all)", "XGB - wage (no zeros)", "XGB - log(wage)"),
  RMSE = c(
    RMSE(xgb_all$pred, test_all$wage),
    RMSE(xgb_pos$pred, test_pos$wage),
    RMSE(xgb_log$pred, test_log$wage)
  ),
  MAE = c(
    MAE(xgb_all$pred, test_all$wage),
    MAE(xgb_pos$pred, test_pos$wage),
    MAE(xgb_log$pred, test_log$wage)
  ),
  R2 = c(
    R2(xgb_all$pred, test_all$wage),
    R2(xgb_pos$pred, test_pos$wage),
    R2(xgb_log$pred, test_log$wage)
  )
)

results_xgb
```
```{r}
importance_xgb <- xgb.importance(
  feature_names = xgb_all$model$feature_names,
  model = xgb_all$model
)

xgb.plot.importance(importance_xgb, top_n = 20)
```

## Neural Network
```{r}

```

# Model Comparison
```{r}
results_all <- rbind(
  results_lm,
  results_tree,
  results_rf,
  results_gbm,
  results_xgb
)

results_all[order(results_all$RMSE), ]
results_all[order(results_all$MAE), ]
results_all[order(-results_all$R2), ]
```

# H20
## AutoML
```{r}
# Installing & initializing H20. In case you have past installations, you should run 
# lines 37-45. If this is the first time installing, you can only run lines 46-47.
if ("package:h2o" %in% search()) { detach("package:h2o", unload=TRUE) }
if ("h2o" %in% rownames(installed.packages())) { remove.packages("h2o") }
pkgs <- c("RCurl","jsonlite")
for (pkg in pkgs) {
  if (! (pkg %in% rownames(installed.packages()))) { install.packages(pkg) }
}
install.packages("h2o", type="source", repos=(c("http://h2o-release.s3.amazonaws.com/h2o/latest_stable_R")))
library(h2o)
```
```{r}
# Call other necessary libraries 
library(data.table)
library(writexl)
library(readr)
```
```{r, message=FALSE, warning=FALSE}
# Initialize H2O cluster
h2o.init()
h2o.no_progress()
```
```{r}
# Set the dataframe as h2o environment 
df <- as.h2o(data_clean)
```
```{r}
# Split dataset into training and validation sets
set.seed(12)
splits <- h2o.splitFrame(df, ratios = c(0.8))
train <- splits[[1]]
valid <- splits[[2]]
```
```{r}
# Set the dependent variable 
dep_var <- "wage"
```
```{r}
automl <- h2o.automl(
  x = setdiff(colnames(df), dep_var), # independent variables
  y = dep_var, # dependent variable
  training_frame = train,
  max_runtime_secs = 600, # maximum time in seconds for AutoML to run
  seed = 12 # set seed for reproducibility
)
```
```{r,message=FALSE, warning=FALSE }
# View leaderboard of models generated by AutoML
lb <- automl@leaderboard
lb_table <- as.data.frame(lb)

knitr::kable(lb_table)

```


# Final Model
```{r}

```

# Our Wage Prediction
```{r}

```

